Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-23 Manuscript/manuscript.Rmd
manuscript edits

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution for the corresponding contact matrix <!--(Figure 1)-->. As an example, for domain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. For the boundary data, in order to accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. That is, for 10 kb resolution data, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was then labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0). The process was repeated for each subsequent resolution.

@@

Using the sets of genomic annotations, we created three types of predictors that were used as features for subsequent models to analyze their respective spatial relationship with TAD boundaries <!--(Figure 3)-->:

@@

To systematically summarize and compare our results, we created an ensemble framework. Each system in the framework was assumed to be composed of a model constructed from a particular cell line, $L = \{GM12878, K562\}$. Within each cell line we evaluated 4 data resolutions, $R = \{10 kb, 25 kb, 50 kb, 100 kb\}$. We then considered 3 separate predictor types as the feature space including $P = \{OC, OP, Distance\}$. The cell line- and resolution-specific data was then split into a 7:3 ratio of training to testing sets. Furthermore, for the task of performance evaluation and prediction, the test set remained constant for all ensemble systems to account for an unbiased comparison. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including $B = \{None, RUS, ROS, SMOTE\}$. Each combination of the above parameters was implemented on chromosome specific sections of the linear genome, and thus, was repeated repeated 21 times (number of autosomal chromosomes, excluding chromosome 9). Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9; see Methods). We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.

@@

We analyzed publically available Hi-C data sets for the GM12878 and K562 cell lines provided by Rao et al. [@rao20143d]. Topologically associating domains were then identified using the Arrowhead algorithm [@durand2016juicebox] obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolutions, for each cell line respectively. We then established a unified approach toward preparing cell line specific TAD boundary data for predictive modelling purposes, irrespective of data resolutions (Figure 1).

@@

Expectedly, the average number of genomic bins, TADs called, and TAD boundaries increased with the increased resolution of cell-line specific Hi-C data (Table 1; Supplementary Tables 3). This correlation with resolution was found to be significant, as evidenced by large intraclass correllation coefficients (ICC) (Table 2; Supplementary Tables 4). However, the class imbalance problem was found to be much more consistant across resolutions in both cell lines. The sparsity of TAD boundaries throughout the genome was found to contribute to heavily imbalanced classes. This imbalance was evidenced by low imbalance ratios, defined as the ratio of the number of genomic bins that overlapped with flanked TAD boundaries ($Y=1$), to the much larger number of genomic bins that did not ($Y=0$). This relationship was not significantly correlated with resoltuion for either cell line, indicating that there existed a severe disparity between classes at all resolutions of Hi-C data.
[??? Very bad and boring. You should tell the STORY, not the report. E.g., "Expectedly, the average number of genomic bins, TADs, and TAD boundaries increased with the increased resolution of Hi-C data (Table 1). However, the imbalance ratio, defined as ..., remained severe (~XXX) at all resolutions (Table 1). In summary, the problem of class imbalance exists at all resolutions of Hi-C data."; !!! editted]

To investigate the relationship between functional genomic annotations and the development of TAD boundaries, we devised 3 different predictor-types that composed the feature space of downstream random forest classification models. These predictor-type specific annotations represented the data matrix, $X = \{X_1, X_2, ..., X_p\}$, where $p$ is the complete set of functional annotations. Overlap counts was the least granular predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 2A). Overlap percents was a more refined predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 2B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 2C).  <!--When measured in basepairs, distances were found to be highly skewed across all resolutions, for both the GM12878 and K562 cell lines (Supplementary Figures 1-2 A-D).--> As a normalization procedure, we performed a $log_{2}$ transformation on the distances prior to modelling (See Methods Section). [??? Introduce all definitions in Results. Put all details in Methods. E.g., distance definition is crucial for understanding results, the fact of log2-transformation is Methods.]

@@

Using the Jaccard index, we were able to determine which group-specific annotations were enriched together on the linear genome. As could be expected, for GM12878, the class of 14 annotations segmenting the chromation into different states exhibited very low pairwise correlations, with a Jaccard index range of (0-0.001) (Table 3). However, specific sets of pairwise histone modifications and transcription factor binding sites (TFBS) were found to be highly enriched. The range of Jaccard indices for the 10 histone modifications was found to be (0.022-0.539). Specifically, H3k27ac, H3k4me1, H3k4me2, and H3k4me3 were found to be highly correlated, high levels of which have been detected surrounding transcription start sites [@barski2007high]. For the 79 TFBS, the range of Jaccard indices was (<0.001-0.425). Here, known mediators of loop extrusion components were found to have high pairwise correlations including CTCF, RAD21, and SMC3 [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Similar profiles of Jaccard indices were exhibited across the sets of annotations for the K562 cell line (Supplementary Table 4). [??? So what? Why did you talk about it?] <!--The high Jaccard indices among select sets of genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation.-->

>>>>

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution for the corresponding contact matrix. As an example, for domain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. For the boundary data, in order to accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. That is, for 10 kb resolution data, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was then labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0). The process was repeated for each subsequent resolution.

@@

Using the sets of genomic annotations, we created three types of predictors that were used as features for subsequent models to analyze their respective spatial relationship with TAD boundaries:

@@

To systematically summarize and compare our results of data driven random forest models, we created an ensemble framework (Figure 1). Each system in the framework was assumed to be composed of a model constructed from a particular cell line, $L = \{GM12878, K562\}$. Within each cell line we evaluated 4 data resolutions, $R = \{10 kb, 25 kb, 50 kb, 100 kb\}$. We then considered 3 separate predictor types as the feature space including $P = \{OC, OP, Distance\}$. The cell line- and resolution-specific data was then split into a 7:3 ratio of training to testing sets. Furthermore, for the task of performance evaluation and prediction, the test set remained constant for all ensemble systems to account for an unbiased comparison. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including $B = \{None, RUS, ROS, SMOTE\}$. Each combination of the above parameters was implemented on chromosome specific sections of the linear genome, and thus, was repeated repeated 21 times (number of autosomal chromosomes, excluding chromosome 9). Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9; see Methods). We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.

@@

We analyzed publically available Hi-C data for the GM12878 and K562 cell lines provided by Rao et al. [@rao20143d]. Topologically associating domains were then identified using the Arrowhead algorithm [@durand2016juicebox] obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolutions, for each cell line respectively. We then established a unified approach toward preparing cell line -specific TAD boundary data for predictive modelling purposes, irrespective of data resolutions (Figure 2).

@@

Expectedly, the median number of genomic bins, TADs called, and TAD boundaries increased with the increased resolution of cell-line specific Hi-C data (Table 1; Supplementary Tables 3). This correlation with resolution was found to be significant, as evidenced by large intraclass correllation coefficients (ICC) (Table 2; Supplementary Tables 4). However, the class imbalance problem was found to be much more consistant across resolutions in both cell lines. The sparsity of TAD boundaries throughout the genome was found to contribute to heavily imbalanced classes. This imbalance was evidenced by low imbalance ratios, defined as the ratio of the number of genomic bins that overlapped with flanked TAD boundaries ($Y=1$), to the much larger number of genomic bins that did not ($Y=0$). This relationship was not significantly correlated with resoltuion for either cell line, indicating that there existed a severe disparity between classes at all resolutions of Hi-C data.
[??? Very bad and boring. You should tell the STORY, not the report. E.g., "Expectedly, the average number of genomic bins, TADs, and TAD boundaries increased with the increased resolution of Hi-C data (Table 1). However, the imbalance ratio, defined as ..., remained severe (~XXX) at all resolutions (Table 1). In summary, the problem of class imbalance exists at all resolutions of Hi-C data."; !!! editted]

@@

The influences of TAD boundaries were assessed using functional genomic annotations, derived by ChIP-seq peak algorithms. Specific groups of annotations were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] and included chromatin state segementations, histone modifications, and transcription factor binding sites (TFBS).

Using the Jaccard index, we found that specific sets of these cell line-specific annotations were highly colocalized together on the linear genome (Table 3). For GM12878, the range of Jaccard indices for the 10 histone modifications was found to be (0.022-0.539). Specifically, H3k27ac, H3k4me1, H3k4me2, and H3k4me3 were found to be highly correlated, high levels of which have been detected surrounding transcription start sites [@barski2007high]. Likewise, for the 79 TFBS, the range of Jaccard indices was (<0.001-0.425). Here, known mediators of loop extrusion components were found to have high pairwise correlations including CTCF, RAD21, and SMC3 [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Similar profiles of Jaccard indices were exhibited across the sets of annotations for the K562 cell line (Supplementary Table 4). [??? So what? Why did you talk about it?; !!!added implication]. The high Jaccard indices among select sets of genomic annotations indicated the colocalization of annotations known to be enriched at TAD boundaries. This gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space of all annotations while retaining the possibly correlated annotations that could be predictive of TAD boundary formation.

We then investigated the relationship between the functional genomic annotations and the development of TAD boundaries by devising 3 different predictor-types that composed the feature space of downstream random forest classification models. These predictor-type specific annotations represented the data matrix, $X = \{X_1, X_2, ..., X_p\}$, where $p$ is the complete set of functional annotations. Overlap counts was the least granular predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 3A). Overlap percents was a more refined predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 3B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 3C).
